A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
omit names

The decision procedure for the ground theory of (co)datatypes $\thD$ determines the
$\thD$-satisfiability of sets of ground constraints over a fixed
signature~$\Sig$. It is formulated as a calculus, whose rules can be applied in
any order until saturation.
%
The rules operate on a finite set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable ${\Varec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = {\Varec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following unfolding rule exhaustively:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}
}
\)
\rn{Unfold}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{examplex}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initially:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

We construct $\Val^\ast$ by exhaustively applying the following rule to $\nf{\Val}$:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} : \tau \in \FV( \Val )
  \quad
  \MU {\Varec{x}}.\; t \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_{\Varec{x}}(\Val)
}{
  \Val := \nf{\Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t \}}
}
\)
\rn{Assign}}
\]
%
The rule chooses an unassigned variable in $\Val$
and assigns it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $\FV( \Val )$ and does not add any variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.

>>>>

The decision procedure for the ground theory of (co)datatypes $\thD$ determines the
$\thD$-satisfiability of sets of ground constraints over a fixed
signature~$\Sig$. It is formulated as a calculus, whose rules are applied
until a contradiction is detected or until saturation.
%
The rules operate on a finite set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

@@

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable ${\Varec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = {\Varec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following unfolding rule exhaustively:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}
}
\)
%\rn{Unfold}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{examplex}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of unfoldings is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initially:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

We construct $\Val^\ast$ by exhaustively applying the following rule to $\nf{\Val}$:
\[
\hbox{\(
\inferrule{
  {\Varec{x}} : \tau \in \FV( \Val )
  \quad
  \MU {\Varec{x}}.\; t \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_{\Varec{x}}(\Val)
}{
  \Val := \nf{\Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t \}}
}
\)
%\rn{Assign}}
\]
%
The rule chooses an unassigned variable in $\Val$
and assigns it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $\FV( \Val )$ and does not add any variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.
